---
title: "ATAC analysis (Sex-biased accessibility)"
author: "Bence Kover"
date: "2026-01-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading dependencies

```{r}
library(DESeq2)
library(limma)
library(edgeR)
library(tidyverse)
library(org.Mm.eg.db)
library(reticulate)
library(Seurat)
```

## Loading datasets

```{r}

version="v_0.02"
h5ad_file <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/pb_h5ad_2026_01_28.h5ad"
# Read the .h5ad file using reticulate
anndata <- reticulate::import("anndata")
py_index <- reticulate::import("pandas")$Index
adata <- anndata$read_h5ad(h5ad_file)

multiome_path = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/pdata_assigned.h5ad"

multiome <- t(anndata$read_h5ad(multiome_path)$X)

multiome_metadata = as.data.frame(anndata$read_h5ad(multiome_path)$obs)

#cell_type_final rename to cell_type, GEO to sample
colnames(multiome_metadata)[colnames(multiome_metadata) == "cell_type_final"] <- "cell_type"
colnames(multiome_metadata)[colnames(multiome_metadata) == "GEO"] <- "sample"

vars_multi = py_to_r(anndata$read_h5ad(multiome_path)$var_names$to_list())

# Convert the data to Seurat object
# Seurat expects the count matrix to be in column-major order (genes x cells)
counts <- t(adata$X) # Transpose to convert to column-major order
meta_data <- as.data.frame(adata$obs) # Convert obs to a data frame
colnames(meta_data)[colnames(meta_data) == "GEO"] <- "sample"
vars <- adata$var
vars <-  py_to_r(adata$var_names$to_list())

#whats the overlap of vars and vars_multi
print(length(intersect(vars, vars_multi)))

# Create the Seurat object
seurat_object <- CreateSeuratObject(counts = counts, meta.data = meta_data)
rownames(seurat_object) <- vars
seurat_object_multi<- CreateSeuratObject(counts = multiome, meta.data = multiome_metadata)
rownames(seurat_object_multi) <- vars_multi

#merge all
all_seurat <- merge(seurat_object, y = c(seurat_object_multi),
                    add.cell.ids = c("all","multi"),
                    project = "atac_pseudobulk")

all_seurat<- JoinLayers(all_seurat)

seurat_object<- all_seurat

#save as /Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/consensus_chromatin_landscape.csv
write.csv(vars, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/consensus_chromatin_landscape.csv", row.names = FALSE)
write.csv(vars, "/Users/k23030440/epitome_code/epitome/data/additional_downloads/consensus_chromatin_landscape/v_0.02/consensus_chromatin_landscape.csv", row.names = FALSE)


root = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/"
figs_folder = paste(root,"Figures/",sep="")

meta_data = seurat_object@meta.data
meta_data$index <- 1:nrow(meta_data)
#rename sample to GEO
meta_data$GEO = meta_data$sample

meta2 = read.csv("/Users/k23030440/epitome_code/epitome/data/curation/v_0.02/cpa.csv")
#rename GEO to sample
colnames(meta2)[colnames(meta2) == "GEO"] <- "sample"

meta2_first <- meta2[!duplicated(meta2$sample), ]

#metadata only keep sample column
meta_data = meta_data[,c("sample","index")]
# Then merge will maintain row dimensions
meta_data <- merge(meta_data, meta2_first, by="sample", all.x=TRUE)
#reorder based on index
meta_data <- meta_data[order(meta_data$index),]

seurat_object <- AddMetaData(seurat_object, meta_data)

#print unique SRA_ID
print(length(unique(meta_data$SRA_ID)))

meta_data %>%
  distinct(SRA_ID, Comp_sex) %>%  
  count(Comp_sex, name = "num_SRA_IDs")

meta_data = seurat_object@meta.data

expression_table = seurat_object@assays$RNA$counts
```

## Filtering object for relevant datasets and performing normalization

```{r}

##########################

library(limma)
library(edgeR)
# Ensure valid column names for assignments and exptype
meta_data$assignments <- make.names(meta_data$cell_type)
meta_data$exptype <- make.names(meta_data$Modality)

# Remove any remaining problematic characters
meta_data$assignments <- gsub("[^A-Za-z0-9_]", "", meta_data$assignments)
meta_data$exptype <- gsub("[^A-Za-z0-9_]", "", meta_data$exptype)

#remove where assignment is Blood
expression_table <- expression_table[,meta_data$assignments != "Erythrocytes"]
meta_data <- meta_data[meta_data$assignments != "Erythrocytes",]

#where Modality is multi_rna turn to multi_atac
meta_data$exptype[meta_data$Modality == "multi_rna"] <- "multi_atac"

# Ensure valid column names for assignments and exptype
meta_data$assignments <- make.names(meta_data$assignments)
meta_data$exptype <- make.names(meta_data$Modality)
meta_data$Sex <- make.names(meta_data$Comp)

# Remove any remaining problematic characters
meta_data$assignments <- gsub("[^A-Za-z0-9_]", "", meta_data$assignments)
meta_data$exptype <- gsub("[^A-Za-z0-9_]", "", meta_data$exptype)
meta_data$Sex <- gsub("[^A-Za-z0-9_]", "", meta_data$Sex)

print(meta_data$exptype)
#add a new column called index
meta_data$index <- 1:nrow(meta_data)
#make index the rownames
rownames(meta_data) <- meta_data$index

# Convert to numeric
meta_data$psbulk_n_cells <- as.numeric(as.character(meta_data$psbulk_n_cells))

expression_table <- sweep(expression_table, 2, meta_data$psbulk_n_cells, "/")

expression_table <- expression_table * median(meta_data$psbulk_n_cells)

#where multi_rna changer to multi_atac
meta_data$exptype[meta_data$exptype == "multi_rna"] <- "multi_atac"


dge = DGEList(counts = expression_table, group = meta_data$assignments)
dge <- calcNormFactors(dge, method = "TMMwsp")
```

# Running limma-voom workflow

```{r}

# Create the design matrix
design <- model.matrix(~ 0 + assignments + assignments:Sex , data = meta_data)

#make name correct with make.names(
colnames(design) <- make.names(colnames(design))
#remove genes expressed in less than 10% of the samples or with less than 1000 total counts
keep_genes <- rowSums(dge$counts > 1) >= 0.01 * ncol(dge$counts) & rowSums(dge$counts) >= 500

#pull out rownames starting with mt or rpl or rps
dge <- dge[keep_genes, ]
dim(dge)


fit <- voom(dge, design, plot=TRUE)

fit <- lmFit(fit, design)

colnames(fit)


#####
#Sex effect
#####
contrast <- makeContrasts(
  assignmentsStem_cells.SexX1,
  levels = design
)

# Compute differential expression
diff_expression <- contrasts.fit(fit, contrast)
diff_expression <- eBayes(diff_expression, robust=TRUE)

# Get top differentially expressed genes
top_genes <- topTable(diff_expression, number = Inf,  adjust.method = "BH", lfc = 1)
top_genes$genes <- rownames(top_genes)
#how many sig
nrow(top_genes[top_genes$adj.P.Val < 0.05, ])
```

# Printing sex-biased genes for stem cells. Then iterating through cell types to obtain sex-biased peaks for each.
```{r}

library(clipr)
# Write the genes to clipboard, one per line
up_genes <- top_genes %>% dplyr::filter(logFC > 0 & adj.P.Val < 0.05)
#clipr::write_clip(paste(up_genes$gene, collapse = "\n"))
down_genes <- top_genes %>% dplyr::filter(logFC < 0 & adj.P.Val < 0.05)
#clipr::write_clip(paste(down_genes$gene, collapse = "\n"))


# Initialize a list to store results
sex_specific_regions <- list()
cell_types <- unique(meta_data$assignments)

# Print column names to debug
print("Design matrix column names:")
print(colnames(design))

results_list <- list()

for (cell_type in cell_types) {
  tryCatch({
    # Create contrast name - be explicit about the interaction term
    contrast_name <- paste0("assignments", cell_type, ".SexX1")
    
    # Check if the contrast exists in the design matrix
    if(!contrast_name %in% colnames(design)) {
      cat(sprintf("Skipping %s: contrast not found in design matrix\n", cell_type))
      next
    }
    
    # Create contrast matrix
    contrast <- makeContrasts(
      sex_effect = contrast_name,
      levels = colnames(design)
    )
    
    # Compute differential expression
    diff_expr <- contrasts.fit(fit, contrast)
    diff_expr <- eBayes(diff_expr, robust=TRUE)
    
    # Get results
    results <- topTable(diff_expr, number=Inf, lfc = 0.5)
    
    # Store significant regions
    male_regions <- rownames(results[results$adj.P.Val < 0.05 & results$logFC > 0,])
    female_regions <- rownames(results[results$adj.P.Val < 0.05 & results$logFC < 0,])
    
    sex_specific_regions[[cell_type]] <- list(
      male = male_regions,
      female = female_regions
    )
    
    cat(sprintf("%s: %d male-specific, %d female-specific regions\n", 
                cell_type, length(male_regions), length(female_regions)))
    
    results$cell_type <- cell_type
    results_list[[cell_type]] <- results
    
  }, error = function(e) {
    cat(sprintf("Error processing %s: %s\n", cell_type, e$message))
  })
}

results_list <- do.call(rbind, results_list)
#split rowname at . to celltype and peak
results_list$cell_type <- sapply(strsplit(rownames(results_list), "\\."), `[`, 1)
results_list$peak <- sapply(strsplit(rownames(results_list), "\\."), `[`, 2)
#keep adj.P.Val < 0.05
results_list <- results_list[results_list$adj.P.Val < 0.05,]
lactotrophs_up <- results_list[results_list$logFC > 0 & results_list$cell_type == "Lactotrophs",]
lactotrophs_down <- results_list[results_list$logFC < 0 & results_list$cell_type == "Lactotrophs",]
gonadotrophs_up <- results_list[results_list$logFC > 0 & results_list$cell_type == "Gonadotrophs",]
gonadotrophs_down <- results_list[results_list$logFC < 0 & results_list$cell_type == "Gonadotrophs",]
```

# Plotting overall number of male-biased and female-biased peaks in each cell type

```{r}
# Create plotting data
plot_data <- data.frame(
  cell_type = names(sex_specific_regions),
  male_regions = sapply(sex_specific_regions, function(x) length(x$male)),
  female_regions = sapply(sex_specific_regions, function(x) length(x$female))
)

# Create long format data
plot_data_long <- rbind(
  data.frame(
    cell_type = plot_data$cell_type,
    count = -plot_data$female_regions,  # Negative for female to show on left
    sex = "Female",
    label = plot_data$female_regions    # Positive label
  ),
  data.frame(
    cell_type = plot_data$cell_type,
    count = plot_data$male_regions,
    sex = "Male",
    label = plot_data$male_regions
  )
)

# Remove empty labels
plot_data_long$label[plot_data_long$label == 0] <- ""

# Add spacing for male labels
plot_data_long$label <- ifelse(plot_data_long$sex == "Male", 
                               paste0("   ", plot_data_long$label),
                               plot_data_long$label)

#where 0 remove
plot_data_long <- plot_data_long[plot_data_long$count != 0, ]

# Calculate total differentially accessible regions per cell type and reorder
plot_data_long <- plot_data_long %>%
  group_by(cell_type) %>%
  mutate(total_diff_acc = sum(abs(count))) %>%
  ungroup() %>%
  arrange(desc(total_diff_acc)) %>%
  mutate(cell_type = factor(cell_type, levels = unique(cell_type)))

# Add fixed label positions
plot_data_long <- plot_data_long %>%
  mutate(label_position = ifelse(count > 0, 5000, -5000))

# Create the plot
plt = ggplot(plot_data_long, aes(x = cell_type, y = count, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = label, y = label_position),
            color = "black",
            size = 5) +
  scale_fill_manual(values = c("Female" = "#FFA500", "Male" = "#63B3ED")) +
  scale_y_continuous(
    labels = abs,
    limits = c(-max(abs(plot_data_long$count)) * 1.2, max(abs(plot_data_long$count)) * 1.2)
  ) +
  coord_flip() +
  labs(
    title = "Sex-biased Chromatin Accessibility Across Cell Types",
    x = "Cell Type",
    y = "Number of Sex-biased Regions",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 16),
    #y tick smaller
    axis.text.y = element_text(size = 14)
  )

# Save the plot
fold = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/"
ggsave(plt, filename = paste0(fold,"sex_biased_accessibility.png"), width = 6, height = 5.5, dpi = 300)
ggsave(plt, filename = paste0(fold,"sex_biased_accessibility.svg"), width = 6, height = 5.5, dpi = 300)

plt
```

# Annotating object for motif enrichment then define function that performs motif enrichment

```{r}
##############
#Signac enrichment
##############
library(BSgenome.Mmusculus.UCSC.mm10)
library(TFBSTools)
library(JASPAR2020)
library(motifmatchr)

library(Signac)
atac_assay <- CreateChromatinAssay(seurat_object[["RNA"]]$counts,sep = c(":", "-"),)

atac_object <- seurat_object
atac_object[["ATAC"]]<- atac_assay
DefaultAssay(atac_object) <- "ATAC"
atac_object[["RNA"]] <- NULL


library(JASPAR2022)

pfm <- getMatrixSet(
  x = JASPAR2022,
  opts = list(collection = "CORE", tax_group = 'vertebrates', all_versions = FALSE)
)

seqlevelsStyle(BSgenome.Mmusculus.UCSC.mm10) <- "UCSC"

library(EnsDb.Mmusculus.v79)
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)

# Change to UCSC style to match the genome and data
seqlevels(annotations) <- paste0('chr', seqlevels(annotations))
genome(annotations) <- "mm10"

# Add the gene information to the ATAC object
Annotation(atac_object) <- annotations

# add motif information
atac_object <- AddMotifs(
  object = atac_object,
  genome = BSgenome.Mmusculus.UCSC.mm10,
  pfm = pfm,
  p.cutoff = 1e-05, w = 10
)

# Function to perform motif analysis for a specific grouping and direction
perform_motif_analysis <- function(sex_specific_regions, grouping_name, direction, output_dir) {
  # Filter for specific grouping and direction
  #grouping_name = "Lactotrophs"
  #direction = "male"
  target_peaks <- sex_specific_regions[[grouping_name]][[direction]]
  if(is.null(target_peaks)) {
    message(sprintf("No peaks found for %s (%s)", grouping_name, direction))
    return(NULL)
  }
  
  #grouping1_up$gene change : to -
  target_peaks <- gsub(":", "-", target_peaks)
  
  
  enriched.motifs <- FindMotifs(
    object = atac_object,
    features = target_peaks,
    background = 100000
  )
  
  closest_features <- ClosestFeature(
    object = atac_object,
    regions = target_peaks
  )
  
  output_file = file.path(output_dir, sprintf("enrichment_results_%s_%s.rds", grouping_name, direction))
  # Save results
  saveRDS(enriched.motifs, output_file)
  write.csv(enriched.motifs, file.path(output_dir, sprintf("enrichment_results_%s_%s.csv", grouping_name, direction)), row.names = FALSE)
  
  #save closest_features
  write.csv(closest_features, file.path(output_dir, sprintf("closest_features_%s_%s.csv", grouping_name, direction)), row.names = FALSE)
  
  return(enriched.motifs)
}
```

# Perform motif analysis on pituitary lineage cell types

```{r}
# Create output directory
output_dir <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/motif_analysis_results/sex/"
dir.create(output_dir, showWarnings = FALSE)

# Get unique groupings from sex_specific_regions$...
groupings <- unique(meta_data$assignments)
groupings <- c("Lactotrophs","Gonadotrophs","Stem_cells","Somatotrophs","Thyrotrophs","Melanotrophs","Corticotrophs")
directions <- c("male", "female")

# Initialize list to store all results
all_enrichment_results <- list()

# Loop through each grouping and direction
for(grouping in groupings) {
  for(direction in directions) {
    message(sprintf("Processing %s (%s)...", grouping, direction))
    
    result <- perform_motif_analysis(sex_specific_regions,
                                     grouping, 
                                     direction, 
                                     output_dir)
    
    if(!is.null(result)) {
      all_enrichment_results[[paste(grouping, direction, sep="_")]] <- result
    }
  }
}

# Save complete results object
saveRDS(all_enrichment_results, file.path(output_dir, "all_motif_results.rds"))

```

# Identify Ar and Esr1 enrichment in sex-biased genes across pituitary cell types

```{r}

# Create summary table
summary_table <- data.frame()

for(name in names(all_enrichment_results)) {
  result <- all_enrichment_results[[name]]
  #keep pvalue < 0.05
  #result <- result[result$p.adjust < 0.05,]
  result$analysis <- name
  #if there are at least 5 enriched motifs add to summary table
  if(nrow(result) > 5) {
    #append to summary table
    summary_table <- rbind(summary_table, result)
    
  }
}

#extract rows where motif.name is Ar or ESR1
summary_table_hormones <- summary_table[summary_table$motif.name %in% c("Ar","ESR1"),]
summary_table_hormones$sex <- gsub("_.*", "", summary_table_hormones$analysis)
summary_table_hormones$cell_type <- gsub(".*_", "", summary_table_hormones$analysis)
summary_table_hormones <- summary_table_hormones %>% mutate(label_position = ifelse(motif.name == "Ar", 5, -5))

#make a plot

# Extract rows where motif.name is Ar or ESR1
summary_table_hormones <- summary_table[summary_table$motif.name %in% c("Ar","ESR1"),]
#rename ESR1 to Esr1
summary_table_hormones$motif.name <- gsub("ESR1", "Esr1", summary_table_hormones$motif.name)

# Extract cell type from analysis column
summary_table_hormones$cell_type <- gsub("_.*", "", summary_table_hormones$analysis)

# Make ESR1 fold enrichment negative for visualization (left side)
summary_table_hormones$fold.enrichment <- ifelse(
  summary_table_hormones$motif.name == "Esr1",
  -summary_table_hormones$fold.enrichment,
  summary_table_hormones$fold.enrichment
)

# Add label position for the motif names
summary_table_hormones <- summary_table_hormones %>% 
  mutate(label_position = ifelse(motif.name == "Ar", 
                                 fold.enrichment + max(fold.enrichment) * 0.1, 
                                 fold.enrichment - max(abs(fold.enrichment)) * 0.1))

# Calculate total enrichment (absolute values) for each cell type for sorting
cell_type_totals <- summary_table_hormones %>%
  group_by(cell_type) %>%
  summarize(total_enrichment = sum(abs(fold.enrichment))) %>%
  arrange(desc(total_enrichment))

# Reorder cell types by total enrichment
summary_table_hormones$cell_type <- factor(
  summary_table_hormones$cell_type,
  levels = cell_type_totals$cell_type
)

#summary_table_hormones
#only keep where pvalue <.05
summary_table_hormones <- summary_table_hormones[summary_table_hormones$pvalue < 0.05,]

# Make a plot
plt <- ggplot(summary_table_hormones, aes(x = cell_type, y = fold.enrichment, fill = motif.name)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = motif.name, y = label_position),
            color = "black",
            size = 5) +
  scale_fill_manual(values = c("Esr1" = "#FFA500", "Ar" = "#63B3ED")) +
  scale_y_continuous(
    labels = abs,
    limits = c(-max(abs(summary_table_hormones$fold.enrichment)) * 1.2, 
               max(abs(summary_table_hormones$fold.enrichment)) * 1.2)
  ) +
  coord_flip() +
  labs(
    title = "Estrogen and Androgen Receptor Motif Enrichment",
    x = "Cell Type",
    y = "Fold Enrichment",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 16)
  )

# Save the plot
fold <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/"
ggsave(plt, filename = paste0(fold, "hormone_receptor_enrichment.png"), width = 4.5, height = 5.5, dpi = 300)
ggsave(plt, filename = paste0(fold, "hormone_receptor_enrichment.svg"), width = 4.5, height = 5.5, dpi = 300)

# Display the plot
plt

# Save summary table
write.csv(summary_table, 
          file.path(output_dir, "motif_analysis_summary.csv"), 
          row.names = FALSE)
```

# Plotting top 20 sex-biased gonadotrophs TF motifs

```{r}
library(tidyverse)
library(patchwork)
library(ggplot2)

# Set the output directory
output_dir <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/motif_analysis_results/sex/"

# Load the Gonadotroph male and female motif lists
gonadotroph_male <- readRDS(file.path(output_dir, "enrichment_results_Gonadotrophs_male.rds"))
gonadotroph_female <- readRDS(file.path(output_dir, "enrichment_results_Gonadotrophs_female.rds"))

# Add a 'sex' column to each data frame
gonadotroph_male$sex <- "male"
gonadotroph_female$sex <- "female"

# Keep only necessary columns before merging
gonadotroph_male <- gonadotroph_male %>%
  dplyr::select(motif.name, pvalue, p.adjust, sex, fold.enrichment)
gonadotroph_female <- gonadotroph_female %>%
  dplyr::select(motif.name, pvalue, p.adjust, sex, fold.enrichment)

# Merge the two data frames
gonadotrophs_merged <- rbind(gonadotroph_male, gonadotroph_female)
#first name is chars before . or :
gonadotrophs_merged$first_name =  gsub("[:.-].*$", "", gonadotrophs_merged$motif.name)
#turn to CAPS
gonadotrophs_merged$first_name = toupper(gonadotrophs_merged$first_name)
#remove duplicates based on first_name keeping the one with the lowest p.adjust
gonadotrophs_merged <- gonadotrophs_merged %>%
  group_by(first_name) %>%
  #keep top 1
  slice_min(order_by = p.adjust, n = 1, with_ties = FALSE
  ) %>%
  ungroup()


# Calculate -log10 p.adjust and filter for p.adjust < 0.05
gonadotrophs_merged <- gonadotrophs_merged %>%
  mutate(p.adjust = p.adjust + 10**-300) %>%
  
  dplyr::filter(p.adjust < 0.05) %>%
  mutate(neg_log10_p_adjust = -log10(p.adjust))

# Find the top 20 motifs based on the maximum -log10 p.adjust value between male and female
top_20_motifs <- gonadotrophs_merged %>%
  group_by(motif.name) %>%
  
  summarize(max_fold.enrichment = max(fold.enrichment, na.rm = TRUE)) %>%
  arrange(desc(max_fold.enrichment )) %>%
  slice_head(n = 30)

# Filter the merged data frame to keep only the top 20 motifs
plot_data <- gonadotrophs_merged %>%
  dplyr::filter(motif.name %in% c(top_20_motifs$motif.name))

# Reorder the motif.name factor based on the average -log10 p.adjust value
plot_data$motif.name <- factor(
  plot_data$motif.name,
  levels = unique(top_20_motifs$motif.name)
)

#if motif.name is NA fill with first_name
motif_names = as.character(plot_data$motif.name)
plot_data$motif.name <- ifelse(is.na(motif_names), plot_data$first_name,motif_names)

#sort levels by fold.enrichment
plot_data <- plot_data %>%
  arrange(desc(fold.enrichment)) %>%
  mutate(motif.name = factor(motif.name, levels = unique(motif.name)))

# Create the plot with male and female bars going in the same direction
plt_gonadotrophs <- ggplot(plot_data, aes(x = fold.enrichment, y = motif.name, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("male" = "#63B3ED", "female" = "#FFA500")) +
  labs(
    title = "Top 20 Enriched Motifs in Gonadotrophs",
    x = "Fold enrichment",
    y = "Motif Name",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 11),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12)
  )

plt_gonadotrophs 

# Save the plot
ggsave(plt_gonadotrophs, filename = file.path(output_dir, "gonadotroph_top20_motifs.png"),
       width = 3.5, height = 5.5, dpi = 300)
ggsave(plt_gonadotrophs, filename = file.path(output_dir, "gonadotroph_top20_motifs.svg"),
       width = 3.5, height = 5.5, dpi = 300)

# Display the plot
print(plt_gonadotrophs)
```

# Plotting top 20 sex-biased lactotrophs TF motifs

```{r}

library(tidyverse)
library(patchwork)
library(ggplot2)

# Set the output directory
output_dir <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/motif_analysis_results/sex/"

# Load the Gonadotroph male and female motif lists
lactotroph_male<- readRDS(file.path(output_dir, "enrichment_results_Lactotrophs_male.rds"))
lactotroph_female <- readRDS(file.path(output_dir, "enrichment_results_Lactotrophs_female.rds"))

# Add a 'sex' column to each data frame
lactotroph_male$sex <- "male"
lactotroph_female$sex <- "female"

# Keep only necessary columns before merging
lactotroph_male <- lactotroph_male %>%
  dplyr::select(motif.name, pvalue, p.adjust, sex, fold.enrichment)
lactotroph_female <- lactotroph_female %>%
  dplyr::select(motif.name, pvalue, p.adjust, sex, fold.enrichment)

# Merge the two data frames
lactotrophs_merged <- rbind(lactotroph_male, lactotroph_female)
#first name is chars before . or :
lactotrophs_merged$first_name =  gsub("[:.-].*$", "", lactotrophs_merged$motif.name)
#turn to CAPS
lactotrophs_merged$first_name = toupper(lactotrophs_merged$first_name)
#remove duplicates based on first_name keeping the one with the lowest p.adjust
lactotrophs_merged <- lactotrophs_merged %>%
  group_by(first_name) %>%
  #keep top 1
  slice_min(order_by = p.adjust, n = 1, with_ties = FALSE
            ) %>%
  ungroup()
  

# Calculate -log10 p.adjust and filter for p.adjust < 0.05
lactotrophs_merged <- lactotrophs_merged %>%
  mutate(p.adjust = p.adjust + 10**-300) %>%
  
  dplyr::filter(p.adjust < 0.05) %>%
  mutate(neg_log10_p_adjust = -log10(p.adjust))

# Find the top 20 motifs based on the maximum -log10 p.adjust value between male and female
top_20_motifs <- lactotrophs_merged %>%
  group_by(motif.name) %>%
  
  summarize(max_fold.enrichment = max(fold.enrichment, na.rm = TRUE)) %>%
  arrange(desc(max_fold.enrichment )) %>%
  slice_head(n = 30)

# Filter the merged data frame to keep only the top 20 motifs
plot_data <- lactotrophs_merged %>%
  dplyr::filter(motif.name %in% c(top_20_motifs$motif.name))

# Reorder the motif.name factor based on the average -log10 p.adjust value
plot_data$motif.name <- factor(
  plot_data$motif.name,
  levels = unique(top_20_motifs$motif.name)
)

#if motif.name is NA fill with first_name
motif_names = as.character(plot_data$motif.name)
plot_data$motif.name <- ifelse(is.na(motif_names), plot_data$first_name,motif_names)

#sort levels by fold.enrichment
plot_data <- plot_data %>%
  arrange(desc(fold.enrichment)) %>%
  mutate(motif.name = factor(motif.name, levels = unique(motif.name)))

# Create the plot with male and female bars going in the same direction
plt_lactotrophs <- ggplot(plot_data, aes(x = fold.enrichment, y = motif.name, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("male" = "#63B3ED", "female" = "#FFA500")) +
  labs(
    title = "Top 20 Enriched Motifs in Lactotrophs",
    x = "Fold enrichment",
    y = "Motif Name",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 11),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12)
  )

plt_lactotrophs 

# Save the plot
ggsave(plt_lactotrophs, filename = file.path(output_dir, "lactotroph_top20_motifs.png"),
       width = 3.5, height = 5.5, dpi = 300)
ggsave(plt_lactotrophs, filename = file.path(output_dir, "lactotroph_top20_motifs.svg"),
       width = 3.5, height = 5.5, dpi = 300)

# Display the plot
print(plt_lactotrophs)
```

# Plotting top 20 sex-biased stem cell TF motifs

```{r}

library(tidyverse)
library(patchwork)
library(ggplot2)

# Set the output directory
output_dir <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE_atac/motif_analysis_results/sex/"

# Load the Gonadotroph male and female motif lists
sc_male <- readRDS(file.path(output_dir, "enrichment_results_Stem_cells_male.rds"))
sc_female <- readRDS(file.path(output_dir, "enrichment_results_Stem_cells_female.rds"))

# Add a 'sex' column to each data frame
sc_male$sex <- "male"
sc_female$sex <- "female"

# Keep only necessary columns before merging
sc_male <- sc_male %>%
  dplyr::select(motif.name, pvalue, p.adjust, sex, fold.enrichment)
sc_female <- sc_female %>%
  dplyr::select(motif.name, pvalue, p.adjust, sex, fold.enrichment)

# Merge the two data frames
scs_merged <- rbind(sc_male, sc_female)
#first name is chars before . or :
scs_merged$first_name =  gsub("[:.-].*$", "", scs_merged$motif.name)
#turn to CAPS
scs_merged$first_name = toupper(scs_merged$first_name)
#remove duplicates based on first_name keeping the one with the lowest p.adjust
scs_merged <- scs_merged %>%
  group_by(first_name) %>%
  #keep top 1
  slice_min(order_by = p.adjust, n = 1, with_ties = FALSE
  ) %>%
  ungroup()


# Calculate -log10 p.adjust and filter for p.adjust < 0.05
scs_merged <- scs_merged %>%
  mutate(p.adjust = p.adjust + 10**-300) %>%
  
  dplyr::filter(p.adjust < 0.05) %>%
  mutate(neg_log10_p_adjust = -log10(p.adjust))

# Find the top 20 motifs based on the maximum -log10 p.adjust value between male and female
top_20_motifs <- scs_merged %>%
  group_by(motif.name) %>%
  
  summarize(max_fold.enrichment = max(fold.enrichment, na.rm = TRUE)) %>%
  arrange(desc(max_fold.enrichment )) %>%
  slice_head(n = 30)

# Filter the merged data frame to keep only the top 20 motifs
plot_data <- scs_merged %>%
  dplyr::filter(motif.name %in% c(top_20_motifs$motif.name))

# Reorder the motif.name factor based on the average -log10 p.adjust value
plot_data$motif.name <- factor(
  plot_data$motif.name,
  levels = unique(top_20_motifs$motif.name)
)

#if motif.name is NA fill with first_name
motif_names = as.character(plot_data$motif.name)
plot_data$motif.name <- ifelse(is.na(motif_names), plot_data$first_name,motif_names)

#sort levels by fold.enrichment
plot_data <- plot_data %>%
  arrange(desc(fold.enrichment)) %>%
  mutate(motif.name = factor(motif.name, levels = unique(motif.name)))

# Create the plot with male and female bars going in the same direction
plt_scs <- ggplot(plot_data, aes(x = fold.enrichment, y = motif.name, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("male" = "#63B3ED", "female" = "#FFA500")) +
  labs(
    title = "Top 20 Enriched Motifs in scs",
    x = "Fold enrichment",
    y = "Motif Name",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 11),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12)
  )

plt_scs 

# Save the plot
ggsave(plt_scs, filename = file.path(output_dir, "sc_top20_motifs.png"),
       width = 3.5, height = 5.5, dpi = 300)
ggsave(plt_scs, filename = file.path(output_dir, "sc_top20_motifs.svg"),
       width = 3.5, height = 5.5, dpi = 300)

# Display the plot
print(plt_scs)
```



