---
title: "RNA pseudobulk analysis (Sex)"
author: "Bence Kover"
date: "2026-01-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading dependencies

```{r}
library(DESeq2)
library(Seurat)
library(tidyverse)
library(reticulate)
library(variancePartition)
```

## Loading datasets

```{r}

version="v_0.02"
# Specify the path to the .h5ad file
h5ad_file <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/pdatas_2026_02_25.h5ad"

# Read the .h5ad file using reticulate
anndata <- reticulate::import("anndata")
py_index <- reticulate::import("pandas")$Index
adata <- anndata$read_h5ad(h5ad_file)

# Convert the data to Seurat object
# Seurat expects the count matrix to be in column-major order (genes x cells)
counts <- t(adata$X) # Transpose to convert to column-major order
meta_data <- as.data.frame(adata$obs) # Convert obs to a data frame
vars <- adata$var
vars <-  py_to_r(adata$var_names$to_list())

# Create the Seurat object
seurat_object <- CreateSeuratObject(counts = counts, meta.data = meta_data)
#print unique Author
rownames(seurat_object) <- vars
seurat_object$assignments <- seurat_object$new_cell_type

root = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/"
figs_folder = paste(root,"Figures/",sep="")

meta_data = seurat_object@meta.data
print(length(unique(meta_data$Author)))
print(length(unique(meta_data$SRA_ID)))

#print sum psbulk_n_cells
print(sum(meta_data$psbulk_n_cells))
expression_table = seurat_object@assays$RNA$counts
unique(meta_data$SRA_ID)
```


## Filtering object for relevant datasets


```{r}

###################################################
### Now with Limma-voom below
library(limma)
library(edgeR)
# Ensure valid column names for assignments and exptype
meta_data$assignments <- make.names(meta_data$assignments)
meta_data$exptype <- make.names(meta_data$Modality)


# Remove any remaining problematic characters
meta_data$assignments <- gsub("[^A-Za-z0-9_]", "", meta_data$assignments)
meta_data$exptype <- gsub("[^A-Za-z0-9_]", "", meta_data$exptype)


#remove where assignment is Blood
expression_table <- expression_table[,meta_data$assignments != "Erythrocytes"]
meta_data <- meta_data[meta_data$assignments != "Erythrocytes",]


meta_data$Age_numeric <- as.numeric(as.character(meta_data$Age_numeric))
meta_data$Comp<- as.numeric(as.character(meta_data$Comp_sex))

#keep only samples between 20 and 150
expression_table <- expression_table[,meta_data$Age_numeric <= 200]
meta_data <- meta_data[meta_data$Age_numeric <= 200,]
expression_table <- expression_table[,meta_data$Age_numeric > 10]
meta_data <- meta_data[meta_data$Age_numeric > 10,]


meta_data$exptype[meta_data$`10X version` == "PARSE_WT_MegaV2"] <- "parse"
meta_data$exptype[meta_data$`10X version` == "Parse_WT"] <- "parse"

#remove those that are parse
expression_table <- expression_table[,meta_data$exptype != "parse"]
meta_data <- meta_data[meta_data$exptype != "parse",]

expression_table <- expression_table[,meta_data$exptype != "multi_rna"]
meta_data <- meta_data[meta_data$exptype != "multi_rna",]

table(meta_data$Modality,meta_data$Comp_sex)

# Ensure valid column names for assignments and exptype
meta_data$assignments <- make.names(meta_data$assignments)
meta_data$exptype <- make.names(meta_data$exptype)
meta_data$Sex <- make.names(meta_data$Comp)


# Remove any remaining problematic characters
meta_data$assignments <- gsub("[^A-Za-z0-9_]", "", meta_data$assignments)
meta_data$exptype <- gsub("[^A-Za-z0-9_]", "", meta_data$exptype)
meta_data$Sex <- gsub("[^A-Za-z0-9_]", "", meta_data$Sex)


#order meta_data based on exptype
#reorder expression_table based on meta_data index
meta_data$sample = colnames(expression_table)


expression_table <- expression_table[,order(meta_data$exptype)]
meta_data <- meta_data[order(meta_data$exptype),]


#print all values
print(meta_data$exptype)
#add a new column called index
meta_data$index <- colnames(expression_table)

```


## Normalization and running voom

```{r}

expression_table <- sweep(expression_table, 2, meta_data$psbulk_n_cells, "/")

expression_table <- expression_table * median(meta_data$psbulk_n_cells)

exp_table_sc = expression_table[,meta_data$exptype == "sc"]
meta_data_sc = meta_data[meta_data$exptype == "sc",]

exp_table_sn = expression_table[,meta_data$exptype == "sn"]
meta_data_sn = meta_data[meta_data$exptype == "sn",]

dge = DGEList(counts = expression_table, group = meta_data$assignments)
# Prepare the DGEList object for limma voom
dge_sc <- DGEList(counts = exp_table_sc, group = meta_data_sc$assignments)
dge_sn <- DGEList(counts = exp_table_sn, group = meta_data_sn$assignments)


library(limma)
library(edgeR)
design <- model.matrix(~0 + assignments +  assignments:Sex, data = meta_data)
colnames(design) <- make.names(colnames(design))
keep_genes <- filterByExpr(dge, design, min.total.count = 300,min.prop= 0.2)
#look at Gpr101
keep_genes[rownames(dge) == "Gpr101"]
print(sum(keep_genes))
dge <- dge[keep_genes, ]
dim(dge)
gene_names <- rownames(dge)
dge_sc <- dge_sc[gene_names,]
dge_sn <- dge_sn[gene_names,]

#norm
dge_sc <- calcNormFactors(dge_sc, method = "TMM")
dge_sn <- calcNormFactors(dge_sn, method = "TMM")



dge$samples = rbind(dge_sc$samples,dge_sn$samples)


meta_data = rbind(meta_data_sc,
                  meta_data_sn
                  )

#set ref level of Sex to X0
meta_data$Sex <- relevel(factor(meta_data$Sex), ref = "X0")
#merge assignment and sex to assignment_sex
meta_data$assignments_sex = paste(meta_data$assignments, meta_data$Sex, sep="_")

# Create the design matrix
design <- model.matrix(~0 + assignments + assignments_sex, data = meta_data)
#from design remove anything with _X0
colnames(design) <- make.names(colnames(design))

fit <- voom(dge, design, plot=TRUE)
```

# Defining contrasts (technically redundant step, but simplifies naming), and defining formula

```{r}

param <- SnowParam(4, "SOCK", progressbar = TRUE)
# The variable to be tested must be a fixed effect
form <- ~0 + assignments + assignments:Sex + (1 | Modality)

L <- makeContrastsDream(form, meta_data, contrasts = c(
  Corticotrophs_sex = "assignmentsCorticotrophs:SexX1 - 0",
  Endothelial_cells_sex = "assignmentsEndothelial_cells:SexX1 - 0",
  Gonadotrophs_sex ="assignmentsGonadotrophs:SexX1 - 0",
  Immune_cells_sex = "assignmentsImmune_cells:SexX1 - 0",
  Lactotrophs_sex = "assignmentsLactotrophs:SexX1 - 0",
  Melanotrophs_sex = "assignmentsMelanotrophs:SexX1 - 0",
  Mesenchymal_cells_sex = "assignmentsMesenchymal_cells:SexX1 - 0",
  Pituicytes_sex = "assignmentsPituicytes:SexX1 - 0",
  Somatotrophs_sex = "assignmentsSomatotrophs:SexX1 - 0",
  Stem_cells_sex =  "assignmentsStem_cells:SexX1 - 0",
  Thyrotrophs_sex = "assignmentsThyrotrophs:SexX1 - 0"))
                                      
# show contrasts matrix
L

# Plot to visualize contrasts matrix
plotContrasts(L)
```

## Fitting dream

```{r}

# estimate weights using linear mixed model of dream
vobjDream <- voomWithDreamWeights(dge, form, meta_data, BPPARAM = param)

fit <- dream(vobjDream, form, meta_data, L=L)
fit <- eBayes(fit)
```

# Examining stem cell hits

```{r}
# Get top differentially expressed genes
top_genes <- topTable(fit, coef = "Stem_cells_sex",
                      number = Inf,lfc=1, adjust.method="BH")
top_genes$genes <- rownames(top_genes)
#how many sig
print(nrow(top_genes %>% filter(adj.P.Val < 0.05 )))

```

# Loading gene annotations to print some TFs. Also save results.

```{r}

cpdb <- read_csv("/Users/k23030440/epitome_code/epitome/data/gene_group_annotation/v_0.01/cpdb.csv")
#keep where category is TF and keep gene
tf_list <- cpdb %>% filter(category == "TF") %>% dplyr::select(gene) %>% distinct()
tf_list <- tf_list$gene
tf_list

print("Up TFs with male")
#rank it by adj.P.val
print(top_genes %>% filter(genes %in% tf_list & logFC > 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val))
print("Down TFs with female")
print(top_genes %>% filter(genes %in% tf_list & logFC < 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val))

library(clipr)
# Write the genes to clipboard, one per line
up_genes <- top_genes %>% filter(logFC > 0 & adj.P.Val < 0.05)
clipr::write_clip(paste(up_genes$gene, collapse = "\n"))
down_genes <- top_genes %>% filter(logFC < 0 & adj.P.Val < 0.05)
clipr::write_clip(paste(down_genes$gene, collapse = "\n"))
background_genes <- rownames(dge)
clipr::write_clip(paste(background_genes, collapse = "\n"))


#save these to /Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_up_genes.csv
write.csv(up_genes, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_sexually_dimorphic_up_genes.csv")
write.csv(down_genes, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_sexually_dimorphic_down_genes.csv")

```

# Extract cell type specific hits

```{r}

# Initialize a list to store results
sex_specific_genes <- list()
cell_types <- unique(meta_data$assignments)

for (cell_type in cell_types) {
  # Create contrast for this cell type's sex effect
  coef =  paste0(cell_type, "_sex")
  # Get results
  results <- topTable(fit, coef=coef,number=Inf,adjust.method="BH", lfc=1)
  
  
  # Store significant genes (positive logFC = male, negative = female)
  male_genes <- rownames(results[results$adj.P.Val < 0.05 & results$logFC > 0,])
  female_genes <- rownames(results[results$adj.P.Val < 0.05 & results$logFC < 0,])
  
  sex_specific_genes[[cell_type]] <- list(
    male = male_genes,
    female = female_genes
  )
  
  cat(sprintf("%s: %d male-specific, %d female-specific genes\n", 
              cell_type, length(male_genes), length(female_genes)))
}

```

# Plotting overall sex-biased genes across cell types

```{r}

# Create dataframe for plotting
plot_data <- data.frame(
  cell_type = names(sex_specific_genes),
  male_genes = sapply(sex_specific_genes, function(x) length(x$male)),
  female_genes = sapply(sex_specific_genes, function(x) length(x$female))
)

# Reshape the data for plotting
plot_data_long <- rbind(
  data.frame(
    cell_type = plot_data$cell_type,
    count = -plot_data$female_genes,  # Negative for female to show on left
    sex = "Female",
    label = plot_data$female_genes    # Positive label
  ),
  data.frame(
    cell_type = plot_data$cell_type,
    count = plot_data$male_genes,
    sex = "Male",
    label = plot_data$male_genes
  )
)

# Create the plot
#where the fill is =, make it ""
plot_data_long$label[plot_data_long$label == 0] <- ""
plot_data_long$label <- ifelse(plot_data_long$sex == "Male", 
                               paste0("   ", plot_data_long$label),
                               plot_data_long$label)

plot_data_long$count <- as.numeric(plot_data_long$count)
# Ensure count is numeric
plot_data_long$count <- as.numeric(plot_data_long$count)

# Precompute vjust for each row
plot_data_long$vjust <- ifelse(plot_data_long$count > 0, 0.6, -0.6)
library(dplyr)

# Ensure count is numeric
plot_data_long$count <- as.numeric(plot_data_long$count)
library(dplyr)

# Calculate total differentially expressed genes per cell type and reorder
plot_data_long <- plot_data_long %>%
  group_by(cell_type) %>%
  mutate(total_diff_exp = sum(abs(count))) %>% # Total absolute differentially expressed genes
  ungroup() %>%
  arrange(desc(total_diff_exp)) %>%            # Order by total genes
  mutate(cell_type = factor(cell_type, levels = unique(cell_type))) # Maintain order

# Add a column for fixed label positions
plot_data_long <- plot_data_long %>%
  mutate(label_position = ifelse(count > 0, 400, -400)) # Labels at fixed positions

# Plot
plt = ggplot(plot_data_long, aes(x = cell_type, y = count, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = label, y = label_position), # Use fixed label positions
            color = "black",
            size = 5) +
  scale_fill_manual(values = c("Female" = "#FFA500", "Male" = "#63B3ED")) +
  scale_y_continuous(
    labels = abs,  # Show absolute values on axis
    limits = c(-max(abs(plot_data_long$count)) * 1.2, max(abs(plot_data_long$count)) * 1.2) # Symmetric axis
  ) +
  coord_flip() +
  labs(
    title = "Sex-biased Genes Across Cell Types",
    x = "Cell Type",
    y = "Number of Sex-biased Genes",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 16),
    #y tick smaller
    axis.text.y = element_text(size = 14)
  )

plt
# Save the plot as PNG and SVG
stem_sex_spec = sex_specific_genes$Stem_cells$female


#save plt as png and svg to /Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1
fold = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/"
ggsave(plt, filename = paste0(fold,"sex_biased_genes.png"), width = 6, height = 5.5, dpi = 300)
ggsave(plt, filename = paste0(fold,"sex_biased_genes.svg"), width = 6, height = 5.5, dpi = 300)

plt

```

# Obtain data frame for all results


```{r}
df <- data.frame()
df_raw <- data.frame()
cell_types <- unique(meta_data$assignments)

for (cell_type in cell_types) {
  # Create contrast for this cell type's sex effect
  coef = paste0(cell_type, "_sex")
  # Get results
  results <- topTable(fit, coef=coef, number=Inf,adjust.method="BH")
  top_genes <- results
  top_genes$cell_type <- gsub("assignments", "", gsub(".SexX1", "", cell_type))
  top_genes$gene <- rownames(top_genes)
  
  #for those where log2fc <0 say female
  top_genes$sex <- ifelse(top_genes$logFC > 0, "male","female")
  top_genes_filtered <- top_genes[top_genes$adj.P.Val < 0.05,]
  
  df = rbind(df, top_genes_filtered)
  df_raw = rbind(df_raw, top_genes)
  
}
  
#add to df a column called "occurs" which is the number of cell types the gene is differentially expressed in
df <- df %>%
  group_by(gene) %>%
  summarize(occurs = n_distinct(cell_type)) %>%
  left_join(df, by = "gene")

df
```

# Saving sex-biased results

```{r}

#write
write.csv(df,"/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sexually_dimorphic_genes.csv")
#only save those with abs log2fc > 1
df_final <- df %>% filter(abs(logFC) > 1)
df_final  <- df_final  %>%
  dplyr::select(-occurs)

df_final <- df_final  %>%
  group_by(gene) %>%
  summarize(occurs = n_distinct(cell_type)) %>%
  left_join(df_final , by = "gene")

df_final 
write.csv(df_final,"/Users/k23030440/epitome_code/epitome/data/sex_dimorphism/v_0.02/sexually_dimorphic_genes.csv")

write.csv(df_raw,"/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sexually_dimorphic_genes_raw.csv")

```

# Keep genes that occur across cell types

```{r}

df_common_female <- df %>%
  filter(sex == "female") %>%
  group_by(gene) %>%
  summarize(n_cell_types = n_distinct(cell_type)) %>%
  filter(n_cell_types >= 3)

# Same for males
df_common_male <- df %>%
  filter(sex == "male") %>%
  group_by(gene) %>%
  summarize(n_cell_types = n_distinct(cell_type)) %>%
  filter(n_cell_types >= 3)


```


# Make a heatmap of genes that occur in the most cell types, but showing this in a balanced way (18 for male bias, 18 for female bias)

```{r}
#####
##### Making heatmap of shared sex genes
#####

df_sex_genes_filtered <- read.csv("/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sexually_dimorphic_genes.csv")

dim(df_sex_genes_filtered )
df_sex_genes_filtered <- df_sex_genes_filtered %>% filter(abs(logFC) > 0)
df_sex_genes_filtered <- df_sex_genes_filtered %>%
  dplyr::select(-occurs)
df_sex_genes_filtered  <- df_sex_genes_filtered   %>%
  group_by(gene) %>%
  summarize(occurs = n_distinct(cell_type)) %>%
  left_join(df_sex_genes_filtered, by = "gene")

df_sex_genes_filtered <-df_sex_genes_filtered %>% filter(occurs > 4)

#keep only those where sign is same direction in all cases
df_sex_genes_filtered <- df_sex_genes_filtered %>%
  group_by(gene) %>%
  filter(all(logFC > 0) | all(logFC < 0)) %>%
  ungroup()

#how many unique genes
print(df_sex_genes_filtered %>% distinct(gene) %>% nrow())

df_sex_genes <- read.csv("/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sexually_dimorphic_genes_raw.csv")
#keep genes in df_sex_filtered
df_sex_genes <- df_sex_genes %>% filter(gene %in% df_sex_genes_filtered$gene)

df_sex_genes <- df_sex_genes %>%
  group_by(gene) %>%
  mutate(avg_log2fc = mean(logFC, na.rm = TRUE)) %>%
  ungroup()

#first 20 gene names
genes_of_interest <- unique(df_sex_genes$gene)

final_results <- df_sex_genes

```



```{r}

# Load required packages
library(dplyr)
library(ggplot2)
library(tidyr)

#only keep those that are avg abs logfc > 1
final_results <- final_results %>%
  group_by(gene) %>%
  mutate(avg_abs_log2fc = mean(abs(logFC), na.rm = TRUE)) %>%
  ungroup()

#remove genes starting with gm or ens !grepl("^Gm|^Gm[0-9]+$|^ENS" !grepl("Rik$"
final_results <- final_results %>% filter(!grepl("^Gm|^ENS|Rik$", gene))


final_results <- final_results %>%
  group_by(gene, sex) %>%
  summarise(
    mean_abs_log2fc = mean(abs(logFC), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(gene) %>%
  slice_max(mean_abs_log2fc, n = 1, with_ties = FALSE) %>%
  rename(majority_sex = sex) %>%
  dplyr::select(gene, majority_sex) %>%
  right_join(final_results, by = "gene")

#keep top 30 genes with highest abs logFC
final_results <- final_results %>% arrange(desc(avg_abs_log2fc))


#assign majority sex to each gene
genes_of_interest1_male <- unique(final_results[final_results$majority_sex == "male",]$gene)[1:18]
genes_of_interest1_female <- unique(final_results[final_results$majority_sex == "female",]$gene)[1:18]
genes_of_interest1 <- c(genes_of_interest1_male, genes_of_interest1_female)

final_results<- final_results %>% filter(gene %in% genes_of_interest1)

final_results <- final_results %>% arrange(desc(avg_log2fc))
genes_of_interest <- unique(final_results$gene)[1:36]

#remove fshb
genes_of_interest <- genes_of_interest[genes_of_interest != "Fshb"]

final_results$logFC <- -final_results$logFC
# Create the heatmap
plot <- final_results %>%
  # Filter to only include genes of interest
  dplyr::filter(gene %in% genes_of_interest) %>%
  # Create a significance flag and preserve input order
  mutate(is_significant = adj.P.Val < 0.05,
         # Add line thickness variable
         line_size = ifelse(gene %in% genes_of_interest, 0.6, 0.4),
         # Convert genes to factor with levels in the exact order specified
         gene = factor(gene, levels = genes_of_interest)) %>%
  # Create the plot
  ggplot(aes(x = cell_type, y = gene, fill = logFC)) +
  geom_tile(aes(color = is_significant, size = line_size), 
            width = 0.8, height = 0.8) +
  
  # Set color scales
  scale_fill_gradient2(low = "#63b3ed", mid = "white", 
                       high = "#ffa500", midpoint = 0) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red"),
                     name = "Significant",
                     labels = c("FALSE" = "Not Significant", "TRUE" = "Significant"),
                     guide = guide_legend(override.aes = list(
                       fill = "white",
                       size = 1,
                       linetype = 1,
                       linewidth = 2
                     ))) +
  scale_size_identity() +
  # Customize theme
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1,
                                   vjust = 1, size = 12),
        axis.title = element_blank(),
        panel.grid = element_blank()) +
  #y axis italics
  theme(axis.text.y = element_text(size = 10, face = "italic")) +
  labs(fill = "logFC")

plot

#ggsave to /Users/k23030440/Library/CloudStorage/OneDrive-King\'sCollegeLondon/PhD/Year_two/Aim\ 1/
ggsave("/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sexually_dimorphic_genes_heatmap.png", 
       plot = plot, 
       width = 6.5, height = 5.5, dpi = 300)

#svg
ggsave("/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sexually_dimorphic_genes_heatmap.svg", 
       plot = plot, 
       width = 6.5, height = 5.5, dpi = 300)

```

# Comparing these new results against v1 results (pre-print version of the atlas)
```{r}
all_results <- read_csv("/Users/k23030440/epitome_code/epitome/data/sex_dimorphism/v_0.01/sexually_dimorphic_genes.csv")
all_results

```

```{r}
all_results <- read_csv("/Users/k23030440/epitome_code/epitome/data/sex_dimorphism/v_0.02/sexually_dimorphic_genes.csv")
table(all_results$cell_type,all_results$sex)
#
all_results_v1 = read_csv("/Users/k23030440/epitome_code/epitome/data/sex_dimorphism/v_0.01/sexually_dimorphic_genes.csv")
table(all_results_v1$cell_type,all_results_v1$sex)

#keep where abs log2fc > 1
all_results <- all_results %>% filter(abs(logFC) > 1)
all_results_v1 <- all_results_v1 %>% filter(abs(logFC) > 1)

library(tidyverse)

# Add version labels
all_results_v2 <- all_results %>%
  mutate(version = "v2")

all_results_v1 <- all_results_v1 %>%
  mutate(version = "v1")

# Keep only relevant columns (adjust if needed)
all_combined <- bind_rows(
  all_results_v1 %>% dplyr::select(cell_type, sex, gene, version),
  all_results_v2 %>% dplyr::select(cell_type, sex, gene, version)
)

# Determine whether gene appears in v1, v2 or both
gene_presence <- all_combined %>%
  distinct(cell_type, sex, gene, version) %>%
  pivot_wider(
    names_from = version,
    values_from = version,
    values_fn = length,
    values_fill = 0
  ) %>%
  mutate(
    category = case_when(
      v1 > 0 & v2 > 0 ~ "both",
      v1 > 0 & v2 == 0 ~ "v1_only",
      v1 == 0 & v2 > 0 ~ "v2_only"
    )
  )

# Count per grouping + direction + category
plot_data <- gene_presence %>%
  count(cell_type, sex,category)

# Split up and down
plot_up <- plot_data %>% filter(sex == "male")
plot_down <- plot_data %>% filter(sex == "female")



ggplot(plot_up, aes(x = cell_type, y = n, fill = category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "both" = "#4CAF50",
    "v1_only" = "#2196F3",
    "v2_only" = "#FF9800"
  )) +
  labs(
    title = "Up-regulated markers",
    y = "Number of genes",
    fill = "Presence"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggplot(plot_down, aes(x =cell_type, y = n, fill = category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "both" = "#4CAF50",
    "v1_only" = "#2196F3",
    "v2_only" = "#FF9800"
  )) +
  labs(
    title = "Down-regulated markers",
    y = "Number of genes",
    fill = "Presence"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r}
library(tidyverse)

# Add version labels
all_results_v2 <- all_results %>%
  mutate(version = "v2")

all_results_v1 <- all_results_v1 %>%
  mutate(version = "v1")

# Combine
all_combined <- bind_rows(
  all_results_v1 %>% dplyr::select(cell_type, sex, gene, adj.P.Val, version),
  all_results_v2 %>% dplyr::select(cell_type, sex, gene, adj.P.Val, version)
)

# Keep genes present in both
both_data <- all_combined %>%
  distinct(cell_type, sex, gene, version, .keep_all = TRUE) %>%
  group_by(cell_type, sex, gene) %>%
  filter(n_distinct(version) == 2) %>%
  ungroup()

# Spread v1 and v2 into columns
both_wide <- both_data %>%
  dplyr::select(cell_type, sex, gene, version, adj.P.Val) %>%
  pivot_wider(
    names_from = version,
    values_from = adj.P.Val
  )


ggplot(both_wide, aes(x = v1, y = v2)) +
  geom_point(aes(color = sex), alpha = 0.2, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "v1 geom_mean_adj_pval (log10)",
    y = "v2 geom_mean_adj_pval (log10)",
    title = "Comparison of geom_mean_adj_pval (log scale)",
    color = "Direction"
  ) +
  theme_bw()

```

```{r}
v1_genes <- all_results_v1 %>%
  distinct(cell_type, sex, gene)

v2_genes <- all_results %>%
  distinct(cell_type, sex, gene)

n_v1 <- nrow(v1_genes)
n_v2 <- nrow(v2_genes)

n_overlap <- inner_join(v1_genes, v2_genes,
                        by = c("cell_type", "sex", "gene")) %>%
  nrow()
pct_v1_in_v2 <- 100 * n_overlap / n_v1
pct_v2_in_v1 <- 100 * n_overlap / n_v2

spearman_cor_log <- cor(
  -log10(both_wide$v1),
  -log10(both_wide$v2),
  method = "spearman",
  use = "complete.obs"
)

spearman_cor_log
cat("Spearman correlation:", round(spearman_cor_log, 3), "\n")
cat("Percent of v1 genes in v2:", round(pct_v1_in_v2, 1), "%\n")
cat("Percent of v2 genes in v1:", round(pct_v2_in_v1, 1), "%\n")

v1_only = anti_join(v1_genes, v2_genes, by = c("cell_type", "sex", "gene"))

```




