---
title: "RNA pseudobulk analysis of age-dependent changes"
author: "Bence Kover"
date: "2026-01-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading dependencies

```{r}
library(DESeq2)
library(Seurat)
library(tidyverse)
library(reticulate)

```

## Loading datasets

```{r}

version="v_0.02"
# Specify the path to the .h5ad file
h5ad_file <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/pdatas_2026_02_25.h5ad"

# Read the .h5ad file using reticulate
anndata <- reticulate::import("anndata")
py_index <- reticulate::import("pandas")$Index
adata <- anndata$read_h5ad(h5ad_file)

# Convert the data to Seurat object
# Seurat expects the count matrix to be in column-major order (genes x cells)
counts <- t(adata$X) # Transpose to convert to column-major order
meta_data <- as.data.frame(adata$obs) # Convert obs to a data frame
vars <- adata$var
vars <-  py_to_r(adata$var_names$to_list())

# Create the Seurat object
seurat_object <- CreateSeuratObject(counts = counts, meta.data = meta_data)
#print unique Author
rownames(seurat_object) <- vars
seurat_object$assignments <- seurat_object$new_cell_type

root = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/"
figs_folder = paste(root,"Figures/",sep="")

meta_data = seurat_object@meta.data
print(length(unique(meta_data$Author)))
print(length(unique(meta_data$SRA_ID)))

#print sum psbulk_n_cells
print(sum(meta_data$psbulk_n_cells))
expression_table = seurat_object@assays$RNA$counts
```

# Proceed with limma-voom analysis

```{r}

###################################################
### Now with Limma-voom below
library(limma)
library(edgeR)
# Ensure valid column names for assignments and exptype
meta_data$assignments <- make.names(meta_data$assignments)
meta_data$exptype <- make.names(meta_data$Modality)

# Remove any remaining problematic characters
meta_data$assignments <- gsub("[^A-Za-z0-9_]", "", meta_data$assignments)
meta_data$exptype <- gsub("[^A-Za-z0-9_]", "", meta_data$exptype)


#remove where assignment is Blood
expression_table <- expression_table[,meta_data$assignments != "Erythrocytes"]
meta_data <- meta_data[meta_data$assignments != "Erythrocytes",]


meta_data$Age_numeric <- as.numeric(as.character(meta_data$Age_numeric))
meta_data$Comp<- as.numeric(as.character(meta_data$Comp_sex))

#shift to 0 and log10
#remove any with values before 0 
expression_table <- expression_table[,meta_data$Age_numeric >= 60]
meta_data <- meta_data[meta_data$Age_numeric >= 60,]

meta_data$Age_numeric <- log10(meta_data$Age_numeric)


# Ensure valid column names for assignments and exptype
meta_data$assignments <- make.names(meta_data$assignments)
meta_data$exptype <- make.names(meta_data$Modality)
meta_data$Sex <- make.names(meta_data$Comp)


# Remove any remaining problematic characters
meta_data$assignments <- gsub("[^A-Za-z0-9_]", "", meta_data$assignments)
meta_data$exptype <- gsub("[^A-Za-z0-9_]", "", meta_data$exptype)
meta_data$Sex <- gsub("[^A-Za-z0-9_]", "", meta_data$Sex)


#add a new column called index
meta_data$index <- 1:nrow(meta_data)
#make index the rownames
rownames(meta_data) <- meta_data$index

expression_table <- sweep(expression_table, 2, meta_data$psbulk_n_cells, "/")

expression_table <- expression_table * median(meta_data$psbulk_n_cells)

#only keeping single-cell experiments
exp_table = expression_table[,meta_data$exptype == "sc"]
meta_data = meta_data[meta_data$exptype == "sc",]

dge = DGEList(counts = exp_table, group = meta_data$assignments)

# Create the design matrix
design <- model.matrix(~0 + assignments +  assignments:Age_numeric , data = meta_data)

#remove genes expressed in less than 10% of the samples or with less than 1000 total counts
keep_genes <-dplyr::filterByExpr(dge, design, min.total.count = 300,min.prop= 0.2)
sum(keep_genes)

dge <- dge[keep_genes, ]
gene_names <- rownames(dge)

#norm
dge <- calcNormFactors(dge, method = "TMM")

colnames(design)<-make.names(colnames(design))

```


# Proceed with limma-voom analysis

```{r}

fit <- voom(dge, design, plot=TRUE)

fit <- lmFit(fit, design)

fit <- eBayes(fit)

```

# Save background set of genes and extract age-dependent genes in stem cells

```{r}

background_genes <- rownames(dge)
write.csv(background_genes, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/enrichment/aging_enrichment/sc_aging_background_genes_2_month.csv", row.names = FALSE)

#####
#Age effect
#####

contrast <- makeContrasts(
  assignmentsStem_cells.Age_numeric,
  levels = design
)

# Compute differential expression
diff_expression <- contrasts.fit(fit, contrast)
diff_expression <- eBayes(diff_expression, robust=TRUE)

# Get top differentially expressed genes
top_genes <- topTable(diff_expression, number = Inf, adjust.method="bonf")
top_genes$genes <- rownames(top_genes)
print(rownames((top_genes)))

#how many sig
print(nrow(top_genes %>% dplyr::filter(adj.P.Val < 0.05 )))

head(top_genes)

cpdb <- read_csv("/Users/k23030440/epitome_code/epitome/data/gene_group_annotation/v_0.01/cpdb.csv")
#keep where category is TF and keep gene
tf_list <- cpdb %>% dplyr::filter(category == "TF") %>% dplyr::select(gene) %>% distinct()
tf_list <- tf_list$gene


print("Up TFs with time")
#rank it by adj.P.val
print(top_genes %>% dplyr::filter(genes %in% tf_list & logFC > 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val))
print("Down TFs with time")
print(top_genes %>% dplyr::filter(genes %in% tf_list & logFC < 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val))


print("Up with time")
#rank it by adj.P.val
print(top_genes %>% dplyr::filter( logFC > 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val))
print("Down  with time")
print(top_genes %>% dplyr::filter( logFC < 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val))



#save these tfs up and down
tfs_up = top_genes %>% dplyr::filter(genes %in% tf_list & logFC > 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val)
tfs_down = top_genes %>% dplyr::filter(genes %in% tf_list & logFC < 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val)

tfs_up

tfs_down
```

# Export changing ligands and receptors and TFs

```{r}

#save to "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/
write.csv(tfs_up, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_tfs_up_2_month.csv")
write.csv(tfs_down, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_tfs_down_2_month.csv")

cpdb <- read_csv("/Users/k23030440/epitome_code/epitome/data/gene_group_annotation/v_0.01/cpdb.csv")
#keep where category is TF and keep gene
ligands  <- cpdb %>%dplyr::filter(category == "ligand") %>% dplyr::select(gene) %>% distinct()
ligands <- ligands$gene
receptors  <- cpdb %>%dplyr::filter(category == "receptor") %>% dplyr::select(gene) %>% distinct()
receptors <- receptors$gene

#ligands /Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/CellChatDBmouse.csv
ligands_up = top_genes %>%dplyr::filter(genes %in% ligands & logFC > 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val)
ligands_down = top_genes %>%dplyr::filter(genes %in% ligands & logFC < 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val)
ligands_up
ligands_down


#receptors
receptors_up = top_genes %>%dplyr::filter(genes %in% receptors & logFC > 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val)
receptors_down = top_genes %>%dplyr::filter(genes %in% receptors & logFC < 0 & adj.P.Val < 0.05) %>% arrange(adj.P.Val)
receptors_up
receptors_down

write.csv(ligands_up, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_ligands_up_2_month.csv")
write.csv(ligands_down, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_ligands_down_2_month.csv")
write.csv(receptors_up, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_receptors_up_2_month.csv")
write.csv(receptors_down, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/sc_receptors_down_2_month.csv")


```

# Extract age-dependent genes for each cell type

```{r}

coefs = c("assignmentsCorticotrophs.Age_numeric",
          "assignmentsEndothelial_cells.Age_numeric", "assignmentsGonadotrophs.Age_numeric",     
          "assignmentsImmune_cells.Age_numeric",      "assignmentsLactotrophs.Age_numeric",      
          "assignmentsMelanotrophs.Age_numeric",      "assignmentsMesenchymal_cells.Age_numeric",
          "assignmentsPituicytes.Age_numeric",        "assignmentsSomatotrophs.Age_numeric",     
          "assignmentsStem_cells.Age_numeric",        "assignmentsThyrotrophs.Age_numeric" )


# Create empty list to store results
all_results <- list()

# Extract cell types from coefficient names
get_cell_type <- function(coef_name) {
  # Remove "assignments" prefix and ".Age_numeric" suffix
  gsub("assignments(.*)\\.Age_numeric", "\\1", coef_name)
}

# Iterate through each coefficient
for(coef in coefs) {
  # Create contrast matrix
  contrast <- makeContrasts(
    contrasts = coef,
    levels = design
  )
  
  # Compute differential expression
  diff_expression <- contrasts.fit(fit, contrast)
  diff_expression <- eBayes(diff_expression, robust=TRUE)
  
  # Get top genes
  top_genes <- topTable(diff_expression, number = Inf,adjust.method="bonf")
  top_genes$genes <- rownames(top_genes)
  #print cell type
  print(get_cell_type(coef))
  #print number of significant genes
  print(nrow(top_genes %>%dplyr::filter(adj.P.Val < 0.05 )))
  
  # Add cell type column
  top_genes$cell_type <- get_cell_type(coef)
  
  # Store results
  all_results[[coef]] <- top_genes
}

# Combine all results into single dataframe
final_results <- do.call(rbind, all_results)

# Reset row names
rownames(final_results) <- NULL

#save to /Users/k23030440/epitome_code/epitome/aging/v_0.02
version = "v_0.02"
write.csv(final_results, paste0("/Users/k23030440/epitome_code/epitome/data/aging/v_0.02/aging_genes_2_month.csv"))


final_results <- read.csv(paste0("/Users/k23030440/epitome_code/epitome/data/aging/v_0.02/aging_genes_2_month.csv"), header = TRUE, stringsAsFactors = FALSE)
```

# Obtain pan-pituitary ageing genes (but specifically for cells of the pituitary lineage, e.g., not mesenchymal cells, pituicytes, immune cells, endothelial cells)

```{r}
#keep only padj < 0.05
final_results_pan_pit <- final_results[final_results$adj.P.Val < 0.05,]
#filter log2fc for abs 0.5
final_results_pan_pit <- final_results_pan_pit[abs(final_results_pan_pit$logFC) >= 0.5,]
#remove where cell_type "Mesenchymal_cells" "Pituicytes"   "Immune_cells" "Endothelial_cells"
final_results_pan_pit <- final_results_pan_pit[!final_results_pan_pit$cell_type %in% c("Mesenchymal_cells", "Pituicytes", "Immune_cells", "Endothelial_cells"),]
#calc avg logFC and geom avg adj pval
final_results_pan_pit$avg_logFC <- ave(final_results_pan_pit$logFC, final_results_pan_pit$genes, FUN = mean)
#geom avg adj pval
final_results_pan_pit$avg_adj_pval <- ave(final_results_pan_pit$adj.P.Val, final_results_pan_pit$genes, FUN = function(x) prod(x)^(1/length(x)))
#add another column saying number of times a given gene occurs
final_results_pan_pit$gene_count <- ave(final_results_pan_pit$genes, final_results_pan_pit$genes, FUN = length)

final_results_pan_pit <- final_results_pan_pit[!duplicated(final_results_pan_pit$genes),]
#at least 3 gene count
final_results_pan_pit <- final_results_pan_pit[final_results_pan_pit$gene_count >= 3,]
final_results_pan_pit

#order according to gene_count
final_results_pan_pit <- final_results_pan_pit[order(final_results_pan_pit$gene_count, decreasing = TRUE),]

#save these genes as csv
#/Users/k23030440/Library/CloudStorage/OneDrive-King\'sCollegeLondon/PhD/Year_two/Aim\ 1/DE/pan_pituitary_ageing.csv
write.csv(final_results_pan_pit, "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/pan_pituitary_ageing_2_month.csv")
#clipboard gene names
unique_pan_pit_genes = final_results_pan_pit$genes
library(clipr)
write_clip(paste(unique_pan_pit_genes, collapse = "\n"))

```

# Across all of the age-dependent pan-pituitary genes, plotting a curated, interpretable subset.

```{r}

# Load required packages
library(dplyr)
library(ggplot2)
library(tidyr)

# Highlight specific genes of interest
genes_of_interest <- c("Wnt10a","Wnt16","Wnt6","Fzd2","Fzd9","Fzd10","Wif1","Lef1","Rspo4","Sfrp1","Sfrp2","Sfrp5","Apcdd1","Tnfrsf19","Sostdc1",
                       "Cdk1","Cdk2","Ccna1","Ccnb1","H19","E2f8","E2f1","Mki67","Top2a","Ube2c","Aurkb",
                       "Lcn2","Cxcl13","Ccl5","H2-Aa","H2-K1","H2-Ab1","Ifit3","Ifit3b","C1qc","C1qa","Il1b","Il6","Il18","Il17rc","Il15ra","Igha")

#keep only where genes in final_results_pan_pit
genes_of_interest <- genes_of_interest[genes_of_interest %in% final_results$genes]
# Create the heatmap
plot <- final_results %>%
  #dplyr::filter to only include genes of interest
 dplyr::filter(genes %in% genes_of_interest) %>%
  # Create a significance flag and preserve input order
  dplyr::mutate(is_significant = adj.P.Val < 0.05,
                # Add line thickness variable
                line_size = ifelse(genes %in% genes_of_interest, 0.6, 0.4),
                # Convert genes to factor with levels in the exact order specified
                genes = factor(genes, levels = genes_of_interest)) %>%
  # Create the plot
  ggplot(aes(x = cell_type, y = genes, fill = logFC)) +
  geom_tile(aes(color = is_significant, size = line_size), 
            width = 0.8, height = 0.8) +
  # Set color scales
  scale_fill_gradient2(low = "#ffa500", mid = "white", 
                       high = "#63b3ed", midpoint = 0) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red"),
                     name = "Significant",
                     labels = c("FALSE" = "Not Significant", "TRUE" = "Significant"),
                     guide = guide_legend(override.aes = list(
                       fill = "white",
                       size = 1,
                       linetype = 1,
                       linewidth = 2
                     ))) +
  scale_size_identity() +
  # Customize theme
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size =10),
        axis.title = element_blank(),
        panel.grid = element_blank()) +
  #y axis italics
  theme(axis.text.y = element_text(size = 10, face = "italic")) +
  labs(fill = "logFC")

plot


#save plot to/Users/k23030440/Library/CloudStorage/OneDrive-King\'sCollegeLondon/PhD/Year_two/Aim\ 1/DE
ggsave(plot, filename = "/Users/k23030440/Library/CloudStorage/OneDrive-King\'sCollegeLondon/PhD/Year_two/Aim\ 1/DE/heatmap_aging_genes_2_month.png", width = 5.1, height = 5.0, dpi = 300)
#also svg
ggsave(plot, filename = "/Users/k23030440/Library/CloudStorage/OneDrive-King\'sCollegeLondon/PhD/Year_two/Aim\ 1/DE/heatmap_aging_genes_2_month.svg", width = 5.1, height = 5.0, dpi = 300)

```

# Plot overall number of age-dependent genes in every cell type

```{r}

# Process the final_results data for plotting
library(ggplot2)
library(dplyr)

# Create a dataframe to store the counts of age-related genes per cell type
age_genes <- list()
cell_types <- unique(final_results$cell_type)

for (cell_type in cell_types) {
  #dplyr::filter results for this cell type
  cell_results <- final_results[final_results$cell_type == cell_type, ]
  
  # Count significant genes (adj.P.Val < 0.05)
  # Positive logFC = increased with age, negative = decreased with age
  increased_genes <- nrow(cell_results[cell_results$adj.P.Val < 0.05 & cell_results$logFC > 0, ])
  decreased_genes <- nrow(cell_results[cell_results$adj.P.Val < 0.05 & cell_results$logFC < 0, ])
  
  age_genes[[cell_type]] <- list(
    increased = increased_genes,
    decreased = decreased_genes
  )
  
  cat(sprintf("%s: %d increased, %d decreased genes with aging\n", 
              cell_type, increased_genes, decreased_genes))
}

# Create dataframe for plotting
plot_data <- data.frame(
  cell_type = names(age_genes),
  increased_genes = sapply(age_genes, function(x) x$increased),
  decreased_genes = sapply(age_genes, function(x) x$decreased)
)
#remove Immune_cells
#plot_data <- plot_data[!grepl("Immune_cells", plot_data$cell_type), ]
# Reshape the data for plotting
plot_data_long <- rbind(
  data.frame(
    cell_type = plot_data$cell_type,
    count = -plot_data$decreased_genes,  # Negative for decreased to show on left
    change = "Decreased",
    label = plot_data$decreased_genes    # Positive label
  ),
  data.frame(
    cell_type = plot_data$cell_type,
    count = plot_data$increased_genes,
    change = "Increased",
    label = plot_data$increased_genes
  )
)

# Format labels
plot_data_long$label[plot_data_long$label == 0] <- ""
plot_data_long$label <- ifelse(plot_data_long$change == "Increased", 
                               paste0("   ", plot_data_long$label),
                               plot_data_long$label)

# Ensure count is numeric
plot_data_long$count <- as.numeric(plot_data_long$count)

# Calculate total differentially expressed genes per cell type and reorder
plot_data_long <- plot_data_long %>%
  group_by(cell_type) %>%
  dplyr::mutate(total_diff_exp = sum(abs(count))) %>% # Total absolute differentially expressed genes
  ungroup() %>%
  arrange(desc(total_diff_exp)) %>%            # Order by total genes
  dplyr::mutate(cell_type = factor(cell_type, levels = unique(cell_type))) # Maintain order

# Add a column for fixed label positions
plot_data_long <- plot_data_long %>%
  dplyr::mutate(label_position = ifelse(count > 0, 400, -400)) # Labels at fixed positions

# Plot
plt = ggplot(plot_data_long, aes(x = cell_type, y = count, fill = change)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = label, y = label_position), # Use fixed label positions
            color = "black",
            size = 5) +
  scale_fill_manual(values = c("Decreased" = "#FFA500", "Increased" = "#63B3ED")) +
  scale_y_continuous(
    labels = abs,  # Show absolute values on axis
    limits = c(-max(abs(plot_data_long$count)) * 1.2, max(abs(plot_data_long$count)) * 1.2) # Symmetric axis
  ) +
  coord_flip() +
  labs(
    title = "Age-related Genes Across Cell Types",
    x = "Cell Type",
    y = "Number of Age-related Genes",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 16)
  )

print(plt)

```

# Save plot

```{r}
# Save the plot as PNG and SVG
output_dir <- "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE"

# PNG with high resolution
ggsave(
  filename = file.path(output_dir, "age_related_genes_2_month.png"),
  plot = plt,
  width = 5.5,
  height = 6,
  dpi = 300
)

# SVG format
ggsave(
  filename = file.path(output_dir, "age_related_genes_2_month.svg"),
  plot = plt,
  width = 5.5,
  height = 6
)

print(plt)
```

# Show how the >2 months results overlap with the full age-dependent gene list. This shows that much fewer genes are picked up and that these mostly increase, not decrease.

```{r}



final_results <- read.csv(paste0("/Users/k23030440/epitome_code/epitome/data/aging/v_0.02/aging_genes_2_month.csv"), header = TRUE, stringsAsFactors = FALSE)
final_results_all_ages <- read.csv(paste0("/Users/k23030440/epitome_code/epitome/data/aging/v_0.02/aging_genes.csv"), header = TRUE, stringsAsFactors = FALSE)
final_results$sig <- ifelse(final_results$adj.P.Val < 0.05, "yes", "no")
final_results_all_ages$sig <- ifelse(final_results_all_ages$adj.P.Val < 0.05, "yes", "no")
final_results$direction <- ifelse(final_results$logFC > 0, "up", "down")
final_results_all_ages$direction <- ifelse(final_results_all_ages$logFC > 0, "up", "down")
table(final_results$cell_type, final_results$sig, final_results$direction)
table(final_results_all_ages$cell_type, final_results_all_ages$sig, final_results_all_ages$direction)




library(dplyr)
library(stringr)
library(ggplot2)

# keep significant genes only
sig_2mo <- final_results %>%
 dplyr::filter(sig == "yes") %>%
  mutate(
    cell_type = str_replace(cell_type, "assignments", ""),
    cell_type = str_replace(cell_type, ":Age_numeric", ""),
    analysis = "Adult ageing"
  )

sig_all <- final_results_all_ages %>%
 dplyr::filter(sig == "yes") %>%
  mutate(
    analysis = "Across the lifecourse"
  )

sig_combined <- full_join(
  sig_2mo,
  sig_all,
  by = c("genes", "cell_type", "direction"),
  suffix = c("_Adult", "_Lifecourse")
) %>%
  mutate(
    category = case_when(
      !is.na(adj.P.Val_Adult) & !is.na(adj.P.Val_Lifecourse) ~ "Both",
      !is.na(adj.P.Val_Adult) &  is.na(adj.P.Val_Lifecourse) ~ "Adult_aging_only",
      is.na(adj.P.Val_Adult) & !is.na(adj.P.Val_Lifecourse) ~ "Lifecourse_only",
    )
  )

plot_df <- sig_combined %>%
  group_by(cell_type, direction, category) %>%
  summarise(n_genes = n(), .groups = "drop")

plot_df$category <- factor(
  plot_df$category,
  levels = c("Lifecourse_only", "Both","Adult_aging_only")
)


plot_stacked <- function(df, dir_label, title_suffix) {
  ggplot(
    df %>%dplyr::filter(direction == dir_label),
    aes(x = cell_type, y = n_genes, fill = category)
  ) +
    geom_bar(stat = "identity") +
    labs(
      x = "Cell type",
      y = "Number of DE genes",
      fill = "Analysis source",
      title = paste(title_suffix, dir_label, "regulated genes")
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold")
    )
}

# Upregulated
plot_up <- plot_stacked(
  plot_df,
  dir_label = "up",
  title_suffix = "Differentially expressed"
)

# Downregulated
plot_down <- plot_stacked(
  plot_df,
  dir_label = "down",
  title_suffix = "Differentially expressed"
)

#ggsave as png
ggsave(
  filename = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/upregulated_genes_comparison.png",
  plot = plot_up,
  width = 5,
  height = 3.5,
  dpi = 300
)

#svg
ggsave(
  filename = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/upregulated_genes_comparison.svg",
  plot = plot_up,
  width = 5,
  height = 3.5
)



ggsave(
  filename = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/downregulated_genes_comparison.png",
  plot = plot_down,
  width = 5,
  height = 3.5,
  dpi = 300
)

#svg
ggsave(
  filename = "/Users/k23030440/Library/CloudStorage/OneDrive-King'sCollegeLondon/PhD/Year_two/Aim 1/DE/downregulated_genes_comparison.svg",
  plot = plot_down,
  width = 5,
  height = 3.5
)


plot_up
plot_down
```





